---
title: "Chakroborty Repoduction"
author: "Gus Howard"
date: "2025-03-7"
output: html_document
---


# Abstract

Chakraborty (2021) examines the relationship between COVID-19 incidence rates and the percentage of people with disabilities across U.S. counties, considering socio-demographic factors like race, ethnicity, poverty status, age, and sex. Using bivariate correlations and generalized estimating equation (GEE) models, the study finds significant positive associations between COVID-19 rates and socially vulnerable populations. This reproduction study aims to verify Chakraborty’s findings for policy, research, and educational purposes by replicating all analyses, including county-level COVID-19 distributions, statistical correlations, and GEE models. The data and code are publicly available on GitHub, ensuring transparency and accessibility for further study.

## Keywords
COVID-19, Race, Disability

## Study Design

This is a reproduction study to understand the links between race/ethnicity and disability and COVID-19 rates.

## Packages

```{r}
packages <- c(
  "tidycensus", "tidyverse", "downloader", "sf", "classInt", "readr",
  "here", "s2", "pastecs", "tmap", "SpatialEpi", "svDialogs",
  "geepack", "knitr", "kableExtra", "foreign", "broom", "dotwhisker", "dotenv"
)

# load and install required packages
if(!require(groundhog)){
  install.packages("groundhog")
  require(groundhog)
}

if(!require(here)){
  install.packages("here")
  require(here)
}

groundhog.day <- "2025-03-08"
set.groundhog.folder(here("data", "scratch", "groundhog"))
groundhog.library(packages, groundhog.day)
```

# DATA

## ACS data
The American Community Survey (ACS) five-year estimate (2014-2018) variables used in the study are outlined in the table below.
Details on ACS data collection can be found at <https://www.census.gov/topics/health/disability/guidance/data-collection-acs.html>

```{r}
 # Query disability demographic data with geographic boundaries

acs <- get_acs(geography = "county",
  table = "S1810",
  year = 2018,
  output = "wide",
  cache_table = TRUE,
  geometry = TRUE,
  keep_geo_vars = TRUE
)

# Query poverty and disability data
acs_pov <- get_acs(geography = "county",
  table = "C18130",
  year = 2018,
  output = "wide",
  cache_table = TRUE
)

# Query state geographic data
state <- get_acs(geography = "state",
  year = 2018,
  variables = c("B01001_001"),
  geometry = TRUE,
  keep_geo_vars = TRUE
)

# Save query results
saveRDS(acs, here("data", "raw", "public", "acs.RDS"))
saveRDS(acs_pov, here("data", "raw", "public", "acs_pov.RDS"))
saveRDS(state, here("data", "raw", "public", "state.RDS"))
```

## Data Transformations

Initial Study does not include Alaska, Hawaii, or Puerto Rico. 
```{r}
# transform coordinate system and fix geometries
acs <- filter(acs, !STATEFP %in% c("02", "15", "72")) %>%
  st_transform(5070) %>%
  st_make_valid()

# Remove Alaska, Hawaii & Puerto Rico,
state <- filter(state, !STATEFP %in% c("02", "15", "72")) %>%
  st_transform(5070)

# Join poverty data to disability data
acs <- acs %>% left_join(acs_pov, by = "GEOID")
rm(acs_pov)
```

```{r Preprocess-ACS}
acs_derived <- mutate(acs,
  dis_pct = S1810_C02_001E / S1810_C01_001E * 100,
  white_pct = S1810_C02_004E / S1810_C01_001E * 100,
  black_pct = S1810_C02_005E / S1810_C01_001E * 100,
  native_pct = S1810_C02_006E / S1810_C01_001E * 100,
  asian_pct = S1810_C02_007E / S1810_C01_001E * 100,
  other_pct =
    (S1810_C02_008E + S1810_C02_009E + S1810_C02_010E) / S1810_C01_001E * 100,
  non_hisp_white_pct = S1810_C02_011E / S1810_C01_001E * 100,
  hisp_pct = S1810_C02_012E / S1810_C01_001E * 100,
  non_hisp_non_white_pct =
    (S1810_C02_001E - S1810_C02_012E - S1810_C02_011E) / S1810_C01_001E * 100,
  bpov_pct = (C18130_004E + C18130_011E + C18130_018E) / C18130_001E * 100,
  apov_pct = (C18130_005E + C18130_012E + C18130_019E) / C18130_001E * 100,
  pct_5_17 = S1810_C02_014E / S1810_C01_001E * 100,
  pct_18_34 = S1810_C02_015E / S1810_C01_001E * 100,
  pct_35_64 = S1810_C02_016E / S1810_C01_001E * 100,
  pct_65_74 = S1810_C02_017E / S1810_C01_001E * 100,
  pct_75 = S1810_C02_018E / S1810_C01_001E * 100,
  male_pct = S1810_C02_002E / S1810_C01_001E * 100,
  female_pct = S1810_C02_003E / S1810_C01_001E * 100
)

# select only relevant geographic identifiers and derived percentages
acs_derived <- acs_derived %>%
  select(
    fips = GEOID,
    statefp = STATEFP,
    county = NAME.x,
    county_st = NAME,
    contains("pct")
  )
```

## COVID Data

```{r}
covid <- read_sf(here("data", "raw", "public", "covidcase080120.gpkg"))

# select and rename the fips code, population, cases, and x,y coordinates
covid <- select(covid,
  fips = FIPS,
  pop = POP_ESTIMA,
  cases = Confirmed,
  x = X, y = Y
)

covid_table <- covid %>%
  mutate(covid_rate = round(covid$cases / covid$pop * 100000, 2)) %>%
  st_drop_geometry()
```

Join COVID and ACS

```{r}
# Join COVID incidence rate data to acs data
acs_covid <- acs_derived %>%
  left_join(covid_table, by = "fips")

# move covid_rate column prior to disability percentages
acs_covid <- acs_covid %>%
  select(fips, statefp, county, county_st, covid_rate, everything())

rm(acs, acs_derived, covid)
```

missing data
```{r}
# county with missing data
filter(acs_covid, is.na(bpov_pct)) %>% st_drop_geometry() %>% kable()

# replace NA with 0
acs_covid[is.na(acs_covid$bpov_pct), ]$bpov_pct <- 0
acs_covid[is.na(acs_covid$apov_pct), ]$apov_pct <- 0
```

COVID incidence Rate Map
```{r}
tm_covid_rates <- tm_shape(acs_covid) +
  tm_polygons("covid_rate",
    title = "COVID-19 Cases per 100,000 people\n(22 January 2020 to 1 August 2020)",
    style = "quantile",
    border.alpha = .2,
    lwd = 0.2,
    palette = "Oranges",
  ) +
  tm_shape(state) +
    tm_borders("grey", lwd = .5) +
  tm_layout(
    legend.position = c("left", "bottom"),
    legend.title.size = 0.8,
    legend.text.size = 0.5
  )

tm_covid_rates
```

Disability Map
```{r map-disability-rates, message = FALSE}
tm_disability_rates <- tm_shape(acs_covid) +
  tm_polygons("dis_pct",
    title = "Percent of People with Disability\n(ACS 2014-2018)",
    style = "quantile",
    border.alpha = .2,
    lwd = 0.2,
    palette = "Oranges"
  ) +
  tm_shape(state) +
  tm_borders("grey", lwd = .5) +
  tm_layout(
    legend.position = c("left", "bottom"),
    legend.title.size = 0.8,
    legend.text.size = 0.5
  )

tm_disability_rates
```

## Summary Stats
```{r}
acs_covid_stats <- acs_covid %>%
  st_drop_geometry() %>%
  select(covid_rate, contains("pct")) %>%
  stat.desc(norm = TRUE) %>%
  round(2) %>%
  t() %>%
  as.data.frame() %>%
  select(min, max, mean, SD = std.dev, ShapiroWilk = normtest.W, p = normtest.p)

acs_covid_stats %>%
  kable(caption = "Reproduced Descriptive Statistics",
        align = "c") %>%
  column_spec(2:6, width_min = "5em") %>%
  column_spec(7, width_min = "2em") %>%
  kable_styling(full_width = FALSE)
```

## Pearsons
```{r pearsons-correlation}
df <- sum(!is.na(acs_covid$dis_pct)) - 2

pearsons_r <- acs_covid %>%
  select(where(is.numeric)) %>%
  st_drop_geometry() %>%
  cor(method = "pearson", use = "everything") %>%
  as.data.frame() %>%
  select(r = covid_rate) %>%
  mutate(
    t = abs(r) / sqrt((1 - r^2) / (df)),
    p = pt(t, df, lower.tail = FALSE)
  ) %>%
  round(3) %>%
  rownames_to_column("variable") %>%
  filter(variable != "covid_rate")

pearsons_r %>%
  kable(caption = "Reproduced Pearson's R",
        align = "c") %>%
  column_spec(2:4, width = "4em") %>%
  kable_styling(full_width = FALSE)
```

```{r compare-pearsons-correlation}
# calculate number of significance stars at p < 0.01 and p < 0.05 levels.
pearsons_r <- mutate(pearsons_r, rp_stars = as.numeric(as.character(cut(p,
  breaks = c(-0.1, 0.01, 0.05, 1),
  labels = c(2, 1, 0)
))))

# join reproduction coefficients to original study coefficients
correlations <- table1 %>%
  filter(variable != "covid_rate") %>%
  select(variable, or_r = r, or_stars = stars) %>%
  left_join(select(pearsons_r, variable, rp_r = r, rp_stars), by = "variable")

# find difference between coefficient and stars
correlations <- correlations %>%
  bind_cols(rename_with(
    correlations[, 4:5] - correlations[, 2:3],
    ~ paste0(.x, "_diff")
  ))

# find coefficients with different directions
correlations <- correlations %>% mutate(rp_dir_diff = (rp_r > 0) - (or_r > 0))

correlations %>%
  kable(caption = "Compare reproduced and original Pearson's R",
        col.names = c("Variable", "R", "Sig. Level", "R", "Sig. Level", "R", "Sig. Level", "Direction"),
        align = "c") %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "Original" = 2, "Reproduced" = 2, "Difference" = 3))
```

```{r original-data-pearson-correlation}
# load author-provided original data
original_gee <- read.csv(here("data", "raw", "public", "chakraborty", "Aug1GEEdata.csv"))

# calculate correlation coefficients using original data
original_gee %>%
  select(Incidence, PerDisable, starts_with("PD")) %>%
  cor(method = "pearson", use = "everything") %>%
  as.data.frame() %>%
  rownames_to_column("or_variable") %>%
  filter(or_variable != "Incidence") %>%
  select(or_variable, unplanned_r = Incidence) %>%
  bind_cols(correlations[, 1:2]) %>%
  mutate(unplanned_r = round(unplanned_r, 3), diff = unplanned_r - or_r) %>%
  select(variable, unplanned_r, or_r, diff) %>%
  kable(caption = "Recalculation of Pearson's R with original data",
        align = "c",
        ) %>%
  kable_styling(full_width = FALSE)
```

```{r compare-incidence-rate}
# recalculate Incidence Rate
original_gee <- original_gee %>%
  mutate(recalc_Incidence = round(Cases / Total_POP * 100000, 2))

# compare recalculation to author-provided original data and print any counties
# with inconsistent results
original_gee %>%
  filter(recalc_Incidence != Incidence) %>%
  select(FIPS = COUNTY_FIPS, State = ST_Name, County = Countyname, Population = Total_POP, Cases, OR_Incidence = Incidence, RPr_Incidence = recalc_Incidence) %>%
  mutate(Difference = RPr_Incidence - OR_Incidence) %>%
  kable(caption = "Counties with inconsistent COVID-19 incidence rate") %>%
  kable_styling(latex_options = "scale_down")
```

```{r join-incidence-rate}
original_incidence <- original_gee %>%
  select(COUNTY_FIPS, or_incidence = Incidence) %>%
  mutate(fips =
           ifelse(COUNTY_FIPS >= 10000,
                  as.character(COUNTY_FIPS),
                  paste0("0", COUNTY_FIPS)
                  )
         )
  # calculates a text version of FIPS code for joining, while adding back
  # the leading '0' if the code was less than 10000

acs_covid <- acs_covid %>%
  left_join(original_incidence, by = "fips")

rm(original_incidence)

acs_covid %>%
  st_drop_geometry %>%
  filter(covid_rate != or_incidence) %>%
  arrange(fips) %>%
  select(county_st, covid_rate, or_incidence) %>%
  kable(caption = "Original incidence rate joined to reproduction data") %>% kable_styling()
```

```{r SpatialEpi-Kulldorff, eval = FALSE, fig.width=4, fig.height=4}
start_time <- Sys.time()
covid_geo <- covid_table %>%
  select(x, y) %>%
  latlong2grid()
# latlong2grid creates approximate equidistant cylindrical grid
# could probably reproject to epsg 5070 and create table with x and y

# calculate expected cases with one strata
expected.cases <- expected(covid_table$pop, covid_table$cases, 1)

# Kulldorff spatial scan statistic
covid_kulldorff <- kulldorff(
  geo = covid_geo,
  cases = covid_table$cases,
  population = covid_table$pop,
  expected.cases = expected.cases,
  pop.upper.bound = 0.5,
  n.simulations = 999,
  alpha.level = 0.05,
  plot = TRUE
)print(
  paste(
    "Run time:",
    round(difftime(Sys.time(), start_time, units = "mins"), 2),
    "minutes"
  ),
  quote = FALSE
)
rm(covid_geo, expected.cases, start_time)

# save results in a file appended with the current date
saveRDS(covid_kulldorff,
  file = here("data", "derived", "public", paste0("covid_kulldorff_", Sys.Date(), ".RDS"))
)
```

Load pre-calculated Kulldorff spatial scan results.
Alternatively, skip or modify this code block to use your own version of the SpatialEpi Kulldorff results.

```{r load-Kulldorff}
# load pre-calculated Kulldorff results
# alternatively, modify the file name with an appended date to load a more current set of results
covid_kulldorff <- readRDS(
  here("data", "derived", "public", "covid_kulldorff.RDS")
)
```

```{r report-Kulldorff}
print("Most likely cluster:", quote = FALSE)
covid_kulldorff$most.likely.cluster
print(
  paste0(
    "Number of Secondary clusters: ",
    length(covid_kulldorff$secondary.clusters)
  ),
  quote = FALSE
)
```

```{r assign-cluster-IDs, message = FALSE}
# list of primary cluster locations (counties)
cluster_locations <- covid_kulldorff$most.likely.cluster$location.IDs.included

# create data frame of clusters and
# calculate the clusterID as the first (center) county FIPS code
clusters <- covid_table[cluster_locations, "fips"] %>%
  mutate(clusterID = covid_table[[cluster_locations[1], "fips"]],
         likelihood = covid_kulldorff$most.likely.cluster$log.likelihood.ratio)

# Get a list of secondary clusters
secondary <- covid_kulldorff$secondary.clusters

# similarly add counties in each secondary cluster to the list of clusters
for (i in secondary) {
  cluster_locations <- i$location.IDs.included
  new_clusters <- covid_table[cluster_locations, "fips"] %>%
    mutate(clusterID = covid_table[[cluster_locations[1], "fips"]],
           likelihood = i$log.likelihood.ratio)
  clusters <- clusters %>% rbind(new_clusters)
}

rm(cluster_locations, secondary, i, new_clusters)
```
